Graph Topological Aspects of Granger Causal Network Learning

### 2019-10-13 presentations/student_dynamical_systems_waterloo_2019/pres.tex
Shortened talk for the CQAM workshop

\def\E{\mathbb{E}}  % Expectation
\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space
\def\R{\mathbb{R}}  % Reals
\def\N{\mathbb{N}}  % Naturals
\def\T{\mathsf{T}}  % Transpose
\def\Z{\mathbb{Z}}  % Transpose
\def\cl{\mathsf{cl}}  % Closure

@@

\date{McGill University, Oct 4 2019}

@@

\begin{frame}{Causality}
  \begin{itemize}
    \item{Do billiard collisions cause motion?}\pause
    \item{Does standing in the street cause cars to stop?}\pause
    \item{Does smoking cause cancer?}\pause
    \item{Does lowering interest rates cause lowering unemployment?}\pause
    \item{Granger causality attemps to address the last question's category.}
  \end{itemize}
\end{frame}

\begin{frame}{Causality in Time Series}
  \begin{itemize}
    \item{Consider $x_t, y_t \overset{\text{i.i.d.}}{\sim} \text{BER}(1/2)$}\pause
    \item{If $z_t = x_{t - 1}$ then since cause preceeds effect, intuitively $x$ causes $z$}\pause
    \item{$z'_t = x_{t - 1} \oplus y_{t - 1}$ is more ambiguous.}\pause
    \item{Issue raised by: \fullcite{transfer_entropy_criticism}} 
  \end{itemize}
\end{frame}

@@

\begin{frame}{Basic Setup}
  \begin{itemize}
    \item{Basic setup is an $n-$dimensional WSS process $x(t)$}\pause
    \item{Wold representation $x(t) = \sum_{\tau = 0}^\infty A(\tau)v(t - \tau)$}\pause
    \item{$\VAR(\infty)$ representation $x(t) = \sum_{\tau = 1}^\infty B(\tau)x(t - \tau) + v(t)$}\pause
    \item{Require $\E[v(t)v(t - \tau)] = \delta_{\tau}\Sigma_v$, with $\Sigma_v$ Diagonal}\pause
  \end{itemize}

  \begin{defn}
    Linear Projection onto Hilbert space $\H \subseteq L_2(\Omega)$
    \begin{equation*}
      \linE{x}{\H} = \underset{u \in \H}{\text{argmin }} ||x - u||
    \end{equation*}

    Minimum Mean Squared Error
    \begin{equation*}
      \xi[x \ |\ \H] = \E (x - \linE{x}{\H})^2      
    \end{equation*}
  \end{defn}
\end{frame}

\begin{frame}{Basic Setup}
  More notation

  \begin{align*}
    \H_t &= \cl \{\sum_{\tau = 0}^p a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, p \in \N\} \subseteq L_2(\Omega)\\
    % H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{align*}\pause

  \begin{itemize}
    \item{$\H_t^{i}$ Hilbert space spanned by $x_i(t)$ alone}
    \item{$\H_t^{-i}$ Hilbert space spanned by $x(t)$ but \textit{excluding} $x_i(t)$}
  \end{itemize}
\end{frame}

\begin{frame}{Granger Causality}
  \begin{defn}
    For the WSS series $x(t)$ we say that component $x_j$
    \textit{Granger-Causes} component $x_i$ (with respect to $x$)
    and write $x_j \gc x_i$ if

    \begin{equation*}
      \linEerr{x_i(t)}{\H_{t - 1}} < \linEerr{x_i(t)}{\H^{-j}_{t - 1}},
    \end{equation*}
  \end{defn}\pause

  \begin{defn}
    We say that $x_j$ pairwise Granger-causes $x_i$ and write
    $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect to
    $(x_i, x_j)$.
  \end{defn}
\end{frame}

@@

  \begin{figure}
    \includegraphics[width=0.75\linewidth]{./figures/var1_example.pdf}
  \end{figure}
\end{frame}

@@

  \centering
  \includegraphics[width=0.75\linewidth]{./figures/var2_example.pdf}
\end{frame}

@@

\begin{frame}{Application Examples}
  Analysis of connectedness of financial institutions: 2002 - 2004

  \begin{columns}
    \begin{column}{0.75\linewidth}
      \includegraphics[width=\linewidth]{./figures/finance_connections2002.png}
    \end{column}
  \end{columns}

  \footfullcite{NBERw16223}
\end{frame}

\begin{frame}{Application Examples}
  Analysis of connectedness of financial institutions: 2006 - 2008

  \begin{columns}
    \begin{column}{0.75\linewidth}
      \includegraphics[width=\linewidth]{./figures/finance_connections2006.png}
    \end{column}
  \end{columns}

  \footfullcite{NBERw16223}
\end{frame}

\begin{frame}{Application Examples}
  Analysis of brain functional connectivity

  \begin{columns}
    \begin{column}{0.55\linewidth}
      \includegraphics[width=\linewidth]{./figures/brain_photo.png}
    \end{column}
  \end{columns}

  \footfullcite{anna_paper2008}
\end{frame}

\begin{frame}{Application Examples}
  Analysis of Gene regulatory networks

@@

\section{Recovery Methods}
\begin{frame}{LASSO Recovery Methods}
  \begin{itemize}
    \item{Small systems ($n = 2$ or $n \lessapprox 10$) with adequate data are amenable to least squares and hypothesis testing}\pause
    \item{Large systems ($n \gtrapprox 20$) or when data is limited are addressed via the LASSO (and variants)}\pause
    \item{Standard linear regression case $y = X\beta^\star + v$:
        \begin{equation*}
          \underset{\beta \in \R^n}{\text{minimize }} \frac{1}{2T}||y - X\beta||_2^2 + \lambda ||\beta||_1
        \end{equation*}
      }\pause
    \item{$\VAR(p)$ models $x(t) = \sum_{\tau = 1}^p B^\star(\tau)x(t - \tau) + v(t)$:
        \begin{equation*}
          \underset{B(\tau) \in \R^{n \times n}}{\text{minimize }} \frac{1}{2T}||x(t) - \sum_{\tau = 1}^p B(\tau)x(t - \tau)||_2^2 + \lambda \sum_{\tau = 1}^p ||B(\tau)||_1
        \end{equation*}
      }
  \end{itemize}
\end{frame}

\begin{frame}{LASSO Recovery Theorems (Linear Regression)}
  \begin{itemize}
    \item{$y = X\beta^\star + v$; $\beta^\star \in \R^n, y \in \R^T$}\pause
    \item{$\widehat{\beta} \in \underset{\beta \in \R^n}{\text{argmin }} \frac{1}{2T}||y - X\beta|| + \lambda ||\widehat{\beta}||_1$}\pause
  \end{itemize}

  \begin{theorem}[Wainwright 2006]
    Suppose $\beta^\star$ has only $s$ non-zero terms:
    $||\beta^\star||_0 = s$, and $v \sim \mathcal{N}(0, \sigma^2)$.
    \pause Then, with high probability, (ignoring some important
    technicalities) there is a constant $C$ s.t. \pause

    \begin{equation*}
      ||\widehat{\beta} - \beta^\star||_2 \le C \sigma \sqrt{\frac{s \mathsf{log } n}{T}}.
    \end{equation*}
  \end{theorem}

  \begin{itemize}
    \item{Roughly, $||\beta^\star - \widehat{\beta}||_2$ is $O\big(\frac{s}{T}\big)$.}
  \end{itemize}
\end{frame}

\begin{frame}{LASSO Recovery Theorems ($\VAR(p)$ case)}
  \begin{itemize}
    \item{$\frac{1}{2T}||x(t) - \sum_{\tau = 1}^p B(\tau)x(t - \tau)||_2^2 + \lambda \sum_{\tau = 1}^p ||B(\tau)||_1$}\pause
  \end{itemize}
  TODO
\end{frame}

\begin{frame}{Structure of Granger Causality Graphs}
  \begin{columns}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_rotation_shape.pdf}\pause
    \end{column}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_star_shaped.pdf}\pause
    \end{column}
  \end{columns}

  \begin{columns}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_loop_shaped.pdf}\pause
    \end{column}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_diamond_shaped.pdf}
    \end{column}
  \end{columns}
\end{frame}
% \item{Confounder: A node $k$ confounds $i, j$ if $k \in \anc{i} \cap \anc{j}$ if exists a
%     path $\gcgpath{k}{i}$ not containing $j$, and a path
%     $\gcgpath{k}{j}$ not containing $i$.}

\section{Pairwise Causality and ``Causal Flow''}
\begin{frame}{Pairwise Testing}
  \begin{itemize}
    \item{Statistical tests for the case $n = 2$ are simple and well established}\pause
    \item{Can we recover all of $\gcg$ by only testing pairwise relations?}\pause
    \item{Thresholding schemes suffer from confounding}\pause
    \item{``Naive'' pairwise testing is so bad, few papers exist on the topic, or it is used as a simple baseline}\pause
    \item{\fullcite{tam2013gene_pwgc}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Necessary Conditions for Pairwise Causality}
  First, when $i, j$ are completely separated:

@@

    \begin{enumerate}
      \item{$j \not\in \anc{i}$ and $i \not\in \anc{j}$}\pause
      \item{$\anc{i}\cap\anc{j} = \emptyset$}\pause
    \end{enumerate}

    then $\H_t^{(i)} \perp \H_t^{(j)}$: $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.
    \pause  Moreover, this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_t^{(i)}} = 0$.\pause
  \end{proposition}

  Note that $i, j$ may still share descendants.
\end{frame}

\begin{frame}{Necessary Conditions for Pairwise Causality}
  $\H_t^{(i)} \perp \H_t^{(j)}$ is a much stronger conclusion that $j \npwgc i$:
  \begin{proposition}
    Consider distinct nodes $i, j$ in a Granger causality graph $\gcg$.
    If

@@

\begin{frame}{Sufficient Conditions for Pairwise Causality}
  \begin{itemize}
    \item{Unfortunately, there is no general converse (i.e. $j \in \pa{i} \centernot\implies j \pwgc i$)}\pause
    \item{We seek special (graph-)\textit{topological} conditions when there is a converse}\pause
  \end{itemize}

@@

\begin{frame}{Strongly Causal Graph Examples}
  \begin{columns}
    \begin{column}{0.33\linewidth}
      \small{Simple Example}
      \includegraphics[width=\linewidth]{../../figures/example_algorithm_step0.pdf}\pause
    \end{column}
    \begin{column}{0.33\linewidth}
      \small{Slightly larger example}
      \includegraphics[width=\linewidth]{../../figures/example_scg.pdf}\pause
    \end{column}
    \begin{column}{0.33\linewidth}
      \small{Similar number of edges, but not strongly causal}
      \includegraphics[width=\linewidth]{../../figures/example_dag.pdf}
    \end{column}
  \end{columns}
\end{frame}

@@

  \begin{itemize}
    \item{Unfortunately, effects of confounding more nuanced}\pause
  \end{itemize}
\end{frame}

\begin{frame}{Persistent Systems}
  \begin{itemize}
    \item{If $k \in \anc{i}\cap\anc{j}$ any of the following may occur:}\pause
    \item{
        $(i \npwgc j$, $j \npwgc i)$,
        $(i \pwgc j$, $j \npwgc i)$,
        $(i \npwgc j$, $j \pwgc i)$,
        $(i \pwgc j$, $j \pwgc i)$}\pause
    \item{Intuitively, $i$ may ``forget'' information from $k$ before the same information reaches $i$.}\pause
  \end{itemize}

  A heuristic ``definition'' (to avoid introducing more cumbersome notation):
  \begin{definition}[Persistent]
    We will say that the process $x(t)$ with Granger causality graph
    $\gcg$ is \textit{persistent} if every node has a self-loop,
    i.e. $\B_{ii}(z) \ne 0$.
  \end{definition}
\end{frame}

\begin{frame}{Persistent Systems}
  Define $H_i(z)$ to be the filter s.t.:

  \begin{equation*}
    H_i(z)x_i(t) = \linE{x_i(t)}{\H_{t - 1}^{(i)}},
  \end{equation*}\pause

  And then with $\sigma_k^2 = \E v_k(t)^2$
  \begin{equation*}
    \label{eqn:T_filter}
    T_{ij}(z) = \sum_{k \in \anc{i} \cap \anc{j}}\sigma_k^2\A_{ik}(z^{-1})(1 - H_i(z^{-1}))(1 - H_j(z))\A_{jk}(z)
  \end{equation*}\pause

  \begin{proposition}
    \label{prop:persistence_converse}
    Fix $i, j \in [n]$ and suppose $\exists k \in \anc{i} \cap \anc{j}$
    which confounds $i, j$.  Then, if $T_{ij}(z)$ is not causal we have
    $j \pwgc i$, and if $T_{ij}(z)$ is not anti-causal we have
    $i \pwgc j$.  Moreover, if $T_{ij}(z)$ is two-sided, then $j \pwgc i \iff i \pwgc j$.
  \end{proposition}
\end{frame}

\section{Pairwise Testing Recovery Methods}
\begin{frame}{Pairwise Recovery}
  \begin{itemize}
    \item{Persistant systems where $T_{ij}(z)$ is not two-sided are highly pathological}\pause
    \item{However, that the following requires an assumption on
        $T_{ij}(z)$ is an unfortunate blemish on the theory as it is
        not a topological requirement}\pause
  \end{itemize}

@@

\begin{frame}{Simulation}
  \begin{itemize}
    \item{Construct random graphs and compare PWGC against LASSO method}\pause
    \item{I obtained the best results with Zou's Adaptive LASSO, but known to suffer somewhat from false positives}
    \item{\fullcite{adaptive_lasso_zou2006}}
  \end{itemize}
\end{frame}

\begin{frame}{Simulation}
  Representative Random Graph Topologies on $n = 50$ Nodes.  Parameter
  $q$ controls the density of edges.
  \begin{columns}
    \begin{column}{0.33\linewidth}
      \begin{figure}
        \caption{SCG}
        \includegraphics[width=\linewidth]{../../figures/example_scg.pdf}
      \end{figure}
    \end{column}\pause
    \begin{column}{0.33\linewidth}
      \begin{figure}
        \caption{DAG $(q = \frac{2}{n})$}
        \includegraphics[width=\linewidth]{../../figures/example_dag.pdf}
      \end{figure}
    \end{column}\pause
    \begin{column}{0.33\linewidth}
      \begin{figure}
        \caption{DAG $(q = \frac{4}{n})$}
        \includegraphics[width=\linewidth]{../../figures/example_dag_q.pdf}
      \end{figure}
    \end{column}\pause
  \end{columns}
\end{frame}

\begin{frame}{Simulation}
  $\VAR(p)$ model with $n = 50, p = 15$ and randomized edge placements.  Left: SCG, Right: DAG ($q = 2/n$)
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \includegraphics[width=0.9\linewidth]{./figures/new_lasso_comparison_scg_pmax15_simulation_mcc_only.png}
    \end{column}

    \begin{column}{0.5\linewidth}
      \includegraphics[width=0.9\linewidth]{./figures/new_lasso_comparison_dag_pmax15_simulation_mcc_only.png}
    \end{column}
  \end{columns}
\end{frame}

@@

>>>>

\def\E{\mathbb{E}}  % Expectation
\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space
\def\Ht{\mathsf{H}}  % Hermitian Transpose
\def\R{\mathbb{R}}  % Reals
\def\N{\mathbb{N}}  % Naturals
\def\T{\mathsf{T}}  % Transpose
\def\Z{\mathbb{Z}}  % Transpose
\def\cl{\mathsf{cl}}  % Closure
\def\ii{\mathsf{i}}  % Imaginary unit

@@

\date{Fields-CQAM Student Workshop on Dynamical Systems and Related Fields, University of Waterloo, Oct 2019}

@@

% \begin{frame}{Causality}
%   \begin{itemize}
%     \item{Do billiard collisions cause motion?}\pause
%     \item{Does standing in the street cause cars to stop?}\pause
%     \item{Does smoking cause cancer?}\pause
%     \item{Does lowering interest rates cause lowering unemployment?}\pause
%     \item{Granger causality attemps to address the last question's category.}
%   \end{itemize}
% \end{frame}

% \begin{frame}{Causality in Time Series}
%   \begin{itemize}
%     \item{Consider $x_t, y_t \overset{\text{i.i.d.}}{\sim} \text{BER}(1/2)$}\pause
%     \item{If $z_t = x_{t - 1}$ then since cause preceeds effect, intuitively $x$ causes $z$}\pause
%     \item{$z'_t = x_{t - 1} \oplus y_{t - 1}$ is more ambiguous.}\pause
%     \item{Issue raised by: \fullcite{transfer_entropy_criticism}}\pause
%     \item{This difficulty is absent for \textit{linear} systems -- the domain of Granger causality}
%   \end{itemize}
% \end{frame}

@@

\begin{frame}{Granger Causality}
  \begin{itemize}
    \item{Basic setup is an $n-$dimensional WSS process $x(t)$}\pause
    \item{$\mathsf{MA}(\infty)$ (Wold) representation $x(t) = \sum_{\tau = 0}^\infty A(\tau)v(t - \tau)$}\pause
    \item{$\VAR(\infty)$ representation $x(t) = \sum_{\tau = 1}^\infty B(\tau)x(t - \tau) + v(t)$}\pause
    \item{Require $\E[v(t)v(t - \tau)] = \delta_{\tau}\Sigma_v$, with $\Sigma_v$ Diagonal}\pause
  \end{itemize}

  \begin{defn}
    For the WSS series $x(t)$ we say that component $x_j$
    \textit{Granger-Causes} component $x_i$ (with respect to $x$)
    and write $x_j \gc x_i$ if

    \begin{equation*}
      \exists \tau > 0 \text{ s.t. } B_{ij}(\tau) \ne 0.
    \end{equation*}
  \end{defn}\pause

  % Need a VAR(2) definition of pairwise causality...
  % \begin{defn}
  %   We say that $x_j$ pairwise Granger-causes $x_i$ and write
  %   $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect to
  %   $(x_i, x_j)$.
  % \end{defn}
\end{frame}

@@

  \begin{figure}
    \includegraphics[width=0.75\linewidth]{../figures/var1_example.pdf}
  \end{figure}
\end{frame}

@@

  \centering
  \includegraphics[width=0.75\linewidth]{../figures/var2_example.pdf}
\end{frame}

@@

% \begin{frame}{Application Examples}
%   Analysis of connectedness of financial institutions: 2002 - 2004

%   \begin{columns}
%     \begin{column}{0.75\linewidth}
%       \includegraphics[width=\linewidth]{../figures/finance_connections2002.png}
%     \end{column}
%   \end{columns}

%   \footfullcite{NBERw16223}
% \end{frame}

% \begin{frame}{Application Examples}
%   Analysis of connectedness of financial institutions: 2006 - 2008

%   \begin{columns}
%     \begin{column}{0.75\linewidth}
%       \includegraphics[width=\linewidth]{../figures/finance_connections2006.png}
%     \end{column}
%   \end{columns}

%   \footfullcite{NBERw16223}
% \end{frame}

% \begin{frame}{Application Examples}
%   Analysis of brain functional connectivity

%   \begin{columns}
%     \begin{column}{0.55\linewidth}
%       \includegraphics[width=\linewidth]{../figures/brain_photo.png}
%     \end{column}
%   \end{columns}

%   \footfullcite{anna_paper2008}
% \end{frame}

\begin{frame}{Application Example}
  Analysis of Gene regulatory networks

@@

\section{Pairwise Causality and ``Causal Flow''}
\begin{frame}{Pairwise Granger Causality}
  \begin{itemize}
    \item{Suppose we observe only $x_i, x_j$ (components of a latent $x$)}\pause
    \item{Test \textit{pairwise} causality: $x_i \pwgc x_j$ (e.g. find
        the best $2-$dimensional model and check for zeros)}\pause
    \item{Statistical tests for the case $n = 2$ are simple and well established}\pause
    \item{Can we recover all of $\gcg$ by only testing pairwise relations?}\pause
    \item{Thresholding schemes suffer from confounding}\pause
    \item{\fullcite{tam2013gene_pwgc}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Necessary Conditions for Pairwise Causality}
  \begin{itemize}
    \item{Interested in relationship between $x_j \gc x_i$ and $x_j \pwgc x_i$}\pause
  \end{itemize}

@@

@@

\begin{frame}{Sufficient Conditions for Pairwise Causality}
  \begin{itemize}
    \item{Unfortunately, there is no general converse (e.g. recall $j \in \pa{i} \centernot\implies j \pwgc i$)}\pause
    \item{We seek special (graph-)\textit{topological} conditions when there is a converse}\pause
  \end{itemize}

@@

\begin{frame}{Strongly Causal Graph Examples}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \small{Simple Example}
      \includegraphics[width=\linewidth]{../../figures/example_algorithm_step0.pdf}\pause
    \end{column}
    \begin{column}{0.5\linewidth}
      \small{Slightly larger example}
      \includegraphics[width=\linewidth]{../../figures/example_scg.pdf}
    \end{column}
  \end{columns}
\end{frame}

@@

  \begin{itemize}
    \item{Unfortunately, effects of confounding more are nuanced}\pause
  \end{itemize}
\end{frame}

% \begin{frame}{Persistent Systems}
%   \begin{itemize}
%     \item{If $k \in \anc{i}\cap\anc{j}$ any of the following may occur:}\pause
%     \item{
%         $(i \npwgc j$, $j \npwgc i)$,
%         $(i \pwgc j$, $j \npwgc i)$,
%         $(i \npwgc j$, $j \pwgc i)$,
%         $(i \pwgc j$, $j \pwgc i)$}\pause
%     \item{Intuitively, $i$ may ``forget'' information from $k$ before the same information reaches $i$.}\pause
%   \end{itemize}

%   \begin{definition}[Persistent]
%     We will say that the process $x(t)$ with Granger causality graph
%     $\gcg$ is \textit{persistent} if every node has a self-loop,
%     i.e. $\B_{ii}(z) \ne 0$.
%   \end{definition}
% \end{frame}

% \begin{frame}{Persistent Systems}
%   Define $H_i(z)$ to be the filter s.t.:

%   \begin{equation*}
%     H_i(z)x_i(t) = \linE{x_i(t)}{\H_{t - 1}^{(i)}},
%   \end{equation*}\pause

%   And then with $\sigma_k^2 = \E v_k(t)^2$
%   \begin{equation*}
%     \label{eqn:T_filter}
%     T_{ij}(z) = \sum_{k \in \anc{i} \cap \anc{j}}\sigma_k^2\A_{ik}(z^{-1})(1 - H_i(z^{-1}))(1 - H_j(z))\A_{jk}(z)
%   \end{equation*}\pause

%   \begin{proposition}
%     \label{prop:persistence_converse}
%     Fix $i, j \in [n]$ and suppose $\exists k \in \anc{i} \cap \anc{j}$
%     which confounds $i, j$.  Then, if $T_{ij}(z)$ is not causal we have
%     $j \pwgc i$, and if $T_{ij}(z)$ is not anti-causal we have
%     $i \pwgc j$.  Moreover, if $T_{ij}(z)$ is two-sided, then $j \pwgc i \iff i \pwgc j$.
%   \end{proposition}
% \end{frame}

\section{Pairwise Testing Recovery Methods}
\begin{frame}{Pairwise Recovery}
  \begin{itemize}
    \item{Additional weak assumptions on $x$ are necessary to avoid pathological cancellation}\pause
    % \item{Persistant systems where $T_{ij}(z)$ is not two-sided are highly pathological}\pause
    \item{However, that the following requires an assumption on
        $T_{ij}(z)$ is an unfortunate blemish on the theory as it is
        not a topological requirement}\pause
  \end{itemize}

@@

\section{Simulations}
\begin{frame}{Simulation}
  \begin{itemize}
    \item{Construct random graphs and compare PWGC against LASSO method}\pause
    \item{I obtained the best competing results with Zou's Adaptive
        LASSO -- but these methods are known to suffer somewhat from
        false positives}
    \item{\fullcite{adaptive_lasso_zou2006}}
  \end{itemize}
\end{frame}

\begin{frame}{Simulation}
  Representative Random Graph Topologies on $n = 50$ Nodes.  Parameter
  $q$ controls the density of edges.
  \begin{columns}
    \begin{column}{0.33\linewidth}
      \begin{figure}
        \caption{SCG}
        \includegraphics[width=\linewidth]{../../figures/example_scg.pdf}
      \end{figure}
    \end{column}\pause
    \begin{column}{0.33\linewidth}
      \begin{figure}
        \caption{DAG $(q = \frac{2}{n})$}
        \includegraphics[width=\linewidth]{../../figures/example_dag.pdf}
      \end{figure}
    \end{column}\pause
    \begin{column}{0.33\linewidth}
      \begin{figure}
        \caption{DAG $(q = \frac{4}{n})$}
        \includegraphics[width=\linewidth]{../../figures/example_dag_q.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Simulation}
  $\VAR(p)$ model with $n = 50, p = 15$ and randomized edge placements.  Left: SCG, Right: DAG ($q = 2/n$)
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \includegraphics[width=0.9\linewidth]{../figures/new_lasso_comparison_scg_pmax15_simulation_mcc_only.png}
    \end{column}

    \begin{column}{0.5\linewidth}
      \includegraphics[width=0.9\linewidth]{../figures/new_lasso_comparison_dag_pmax15_simulation_mcc_only.png}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{Conclusion}
\begin{frame}{Conclusion}
  \begin{itemize}
    \item{The topology of $\gcg$, all else equal, may effect LASSO recovery}\pause
    \item{Granger causality suffers from some pathologies with ``causal flow''}\pause
    \item{$\gcg$ can be recovered from pairwise tests if it is strongly causal}\pause
    \item{Pairwise recovery algorithm performs surprisingly well, even if $\gcg$ is not strongly causal.}\pause
    \item{QUESTIONS?}
  \end{itemize}
\end{frame}

### 2019-10-13 presentations/student_dynamical_systems_waterloo_2019/rkinnear_abstract.tex
Shortened talk for the CQAM workshop

% \usepackage{graphicx}
% \usepackage{caption}
\usepackage{subcaption}
\newtheorem{theorem}{Theorem}

@@

A substantial portion of the work on this topic is devoted to
generalizing the methods of Granger to nonlinear systems.  For
example, linear Granger causality is a special case of the information
theoretic notion of transfer entropy \cite{barnett2009granger}.
However, like many attempts at quantifying causation, transfer entropy
suffers from some difficulties of interpretation
\cite{transfer_entropy_criticism}.  Although the criticisms raised by
\cite{transfer_entropy_criticism} do not arise in the case of linear
Granger causality, other pathologies are known.  Our work sheds light
on some of the challenges that remain in the linear case, and in
particular focuses on topological particularities of the underlying
network of interactions.

In general, we are presented with a multitude of jointly wide sense
stationary time series $x_1(t), \ldots, x_n(t)$ and seek to construct
a (Granger-)causal graph $\gcg$ where there is an edge
$(j, i) \in \gcg$ whenever $x_j$ Granger causes $x_i$ after
conditioning on each $k \not\in \{i, j\}$.  Depending on the
interpretation or on the application, the graph $\gcg$ captures the
flows of ``information'' or ``energy'' throughout a network of
interacting processes.  However, it may be the case that we observe
$x_j$ and $x_i$ alone -- in such cases, simple examples establish that
even if $x_j$ Granger causes $x_i$ conditioned on the rest of the
observations, it need not be that $x_j$ Granger causes $x_i$ pairwise
(i.e. without conditioning).

Pathologies of pairwise causation clash with our intuition that causal
relations in $\gcg$ should be transitive, that is, that if there is
some ``true'' (conditional) causation path
$j \rightarrow a_0 \rightarrow \cdots \rightarrow a_r \rightarrow i$
in $\gcg$ then it should hold that $x_j$ pairwise causes $x_i$.  While
this does not hold in general, our work analyzes a particular type of
graph topology, which we deem a ``strongly causal'' graph, wherein
natural intuitions about causal flow \textit{do} hold.  Moreover, we
prove that (with modest additional technical assumptions) that in such
a graph the complete conditional causation graph $\gcg$ can be
recovered by testing only pairwise interactions of nodes.  That is,
the complete causal structure of $\gcg$ can be recovered through
simple one way unconditional causality tests.

@@

% \begin{figure}[h]
%   \centering
%   \begin{subfigure}[b]{0.45\textwidth}
%     \includegraphics[width=\linewidth]{../../figures/example_algorithm.pdf}
%   \end{subfigure}
%   \begin{subfigure}[b]{0.45\textwidth}
%     \includegraphics[width=\linewidth]{../../figures/example_algorithm2.pdf}
%   \end{subfigure}
% \end{figure}

\begin{figure}[h]
  \centering
  \caption{Recovering Large Networks via Pairwise Causality Tests}
  \label{fig:large_n_figure}
  \includegraphics[width=0.7\linewidth]{../../figures/new_increasing_n_simulation.pdf}
\end{figure}

Our concept of a strongly causal graph topology initially arose
artificially as a technical condition to understand the intuition of
causality flow in $\gcg$ and to establish Theorem
\ref{thm:scg_recovery}.  However, examples from the gene regulatory
network literature suggest that the dynamics of some natural
interacting systems may have an underlying interaction graph which is
in fact strongly causal, or at least nearly so.  Simulation evidence
also suggests that our algorithm for graph recovery via pairwise
testing can outperform the widely employed LASSO (and variations
thereon) for systems which satisfy (or nearly satisfy) the strongly
causal condition; in the process we make the incidental observation
that, all else equal, the LASSO exhibits superior performance on the
strongly causal topology than on a general directed acyclic graph,
suggesting that topological aspects of the underlying causality graph
may be of interest as an aspect unique to time series data in
comparison to ordinary regression.  Finally, the results of Figure
\ref{fig:large_n_figure} show that when topological assumptions are
met, it is possible to accurately identify extremely large networks
($n > 1000$, implying more than $10^6$ possible edges), and that the
underlying graph topology is an important determination of estimation
accuracy in addition to the simple size and sparsity of the network.

>>>>

\newtheorem{theorem}{Theorem}

@@

A substantial portion of the work on this topic is devoted to
generalizing the methods of Granger to nonlinear systems.  For
example, linear Granger causality is a special case of the information
theoretic notion of transfer entropy \cite{barnett2009granger}.
However, like many attempts at quantifying causation, transfer entropy
suffers from some difficulties of interpretation
\cite{transfer_entropy_criticism}.  Although the criticisms raised by
\cite{transfer_entropy_criticism} do not arise in the case of linear
Granger causality, other pathologies are known and our work sheds
light on some of the challenges that remain in the linear case.  In
particular, we focus on topological particularities of the underlying
network of interactions, asking: what class of network topologies are
compatible with our intuitions about causation?

In general, we are presented with jointly wide sense stationary
processes $x_1(t), \ldots, x_n(t)$ and seek to construct a Granger
causal graph $\gcg$ where there is an edge $(j, i) \in \gcg$ whenever
$x_j$ Granger causes $x_i$ conditional on each $k \not\in \{i, j\}$.
Depending on the interpretation or on the application, the graph
$\gcg$ captures the flows of ``information'' or ``energy'' throughout
a network of interacting processes.  However, it may be the case that
we observe $x_j$ and $x_i$ alone -- in such cases, simple examples
establish that even if $x_j$ Granger causes $x_i$ conditioned on the
rest of the observations, it need not be that $x_j$ Granger causes
$x_i$ pairwise (i.e. without conditioning).

Pathologies of pairwise causation clash with our intuition that, for
example, causal relations in $\gcg$ should be transitive: if there is
some ``true'' (conditional) causation path
$j \rightarrow a_0 \rightarrow \cdots \rightarrow a_r \rightarrow i$
in $\gcg$ then it should hold that $x_j$ pairwise causes $x_i$.  While
this does not hold in general, our work analyzes a particular type of
graph topology, which we deem a ``strongly causal'' graph, wherein
natural intuitions about causal flow \textit{do} hold.  Moreover, we
prove that (with modest additional technical assumptions) that in such
a graph the complete conditional causation graph $\gcg$ can be
recovered by testing only pairwise interactions of nodes.  That is,
the complete causal structure of $\gcg$ can be recovered through
simple one way unconditional causality tests.

@@

\begin{figure}[h]
  \centering
  \caption{Recovering Large Networks via Pairwise Causality Tests}
  \label{fig:large_n_figure}
  \includegraphics[width=0.6\linewidth]{abstract_figure.pdf}
\end{figure}

Our concept of a strongly causal graph topology initially arose
artificially as a technical condition to understand the intuition of
causality flow in $\gcg$ and to establish Theorem
\ref{thm:scg_recovery}.  However, examples from the gene regulatory
network literature suggest that the dynamics of some natural
interacting systems may have an underlying interaction graph which is
in fact strongly causal, or at least nearly so.  Simulation evidence
also suggests that our algorithm for graph recovery via pairwise
testing can outperform the widely employed LASSO (and variations
thereon) for systems which satisfy (or nearly satisfy) the strong
causality condition; in the process we make the incidental observation
that, all else equal, the LASSO exhibits marginally superior
performance on the strongly causal topology than on a general directed
acyclic graph, suggesting that topological aspects of the underlying
causality graph may be of interest as an aspect unique to time series
data in comparison to ordinary regression.  Finally, the results of
Figure \ref{fig:large_n_figure} show that when topological assumptions
are met, it is possible to accurately identify extremely large
networks ($n > 1000$, implying more than $10^6$ possible edges), and
that the underlying graph topology is an important determination of
estimation accuracy in addition to the simple size and sparsity of the
network.
